Image Caption Generation With CLIP+GPT-2 Model

### 2025-05-04 Latex_Paper/Paper.tex
update all

Motivated by this progress, we focus our project on replicating a CLIP+GPT-2-based image captioning system, with inspriation from Nukrai et al.\cite{Nukrai2022}. Through this project, we seek to explore and deepen our understanding of 
machine learning, as well as its broader applications in fields such as computer vision and natural language generation.
\begin{itemize}
    \item Leverage the pretrained CLIP model to extract semantically rich image embeddings without the need for training a custom vision encoder.
    \item Utilize the generative capabilities of GPT-2 to produce fluent and coherent natural language captions.
    \item Bridge the gap between visual and textual modalities by introducing a projection layer that maps image embeddings into GPT-2's input space.
    \item Enable flexible and data-efficient image captioning, where the visual semantics guide the generation through prefix-based conditioning.
    \item Evaluate the quality of the generated captions using standard metrics such as BLEU and CIDEr, in order to quantitatively assess the model's accuracy and relevance.
\end{itemize}
\pagebreak
\section{Background and Related Work}

@@

We compare these classical approaches with SimCLR to understand its advantages and limitations.
\pagebreak
\section{Methodology}

\subsection{Implementation Details}
\textbf{Backbone Network:} We use a ResNet-18 as the feature encoder. The final representation is fed into
a two-layer MLP projection head.

@@

\textbf{Training Details:} We train SimCLR on CIFAR-10 and STL-10 using:
\begin{itemize}
  \item Batch size: \{128, 256, 512\}
  \item Temperature parameter: \{0.07, 0.1, 0.5\}
  \item Optimizer: Adam with a learning rate of $3 \times 10^{-4}$
  \item Number of epochs: 200
\end{itemize}

>>>>

Motivated by this progress, we focus our project on replicating a CLIP+GPT-2-based image captioning system, with inspriation from Nukrai et al.\cite{Nukrai2022}. Through this project, we seek to explore and deepen our understanding of 
machine learning, as well as its broader applications in fields such as computer vision and natural language generation.

Our goal is to:
\begin{itemize}
    \item Leverage the pretrained CLIP model to extract semantically rich image embeddings without the need for training a custom vision encoder.
    \item Utilize the generative capabilities of GPT-2 to produce fluent and coherent natural language captions.
    \item Bridge the gap between visual and textual modalities by introducing a projection layer that maps image embeddings into GPT-2's input space.
    \item Enable flexible and data-efficient image captioning, where the visual semantics guide the generation through prefix-based conditioning.
    \item Evaluate the quality of the generated captions using standard metrics such as BLEU and CIDEr, in order to quantitatively assess the model's accuracy and relevance.
\end{itemize}
\pagebreak
\section{Background and Related Work}

@@

We compare these advanced approaches of image caption generation to understand its 
advantages and limitations, and implement these methods in the process of model training to get our own model.
\pagebreak
\section{Methodology}

\subsection{Implementation Details}
\textbf{Detail of Implementation and pipeline:} This is the pipeline of 
our entire training process, which mainly includes multiple steps: 
image input, feature extraction and encoding, feature processing, 
feature decoding, model training and error evaluation. We will also 
introduce the optimization we used.

@@

\textbf{Training Details:} We train our model using these hyperparameter after cross-validation:
\begin{itemize}
  \item Prefix length: 40
  \item Transformer layers: 8
  \item Optimizer: Adam with a learning rate of $5 \times 10^{-5}$
  \item Number of epochs: 20
  \item Noise rate: 10\%
\end{itemize}
